### 2014-05-19 patchy-selection-paper.tex
edited up until applications

%%%%%%%%%%%%
\subsection{Multiple patches}
We can extend this reasoning to multiple patches of unadapted habitat 
in which the $B$ allele can facilitate local adaptation. 
Under our assumptions of mutational exclusion and rapid establishment, 
each patch will adapt through establishment of a single copy of the $B$ allele, 
either by migration or mutation.
If we imagine that the times between successive establishments is the result of many nearly-independent attempts
with small probability of success,
the process of occupation is well approximated by a discrete-state Markov chain,
whose state is the labeling of patches according to the colonizing allele
(or none, for those that haven't yet adapted).

If not yet adapted, a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$.
Without loss of generality, assume that patches $1, \ldots, k$ are already adapted,
and these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch, 
then the total rate of adaptation through migration is approximately
\begin{equation}
  \frac{ s_p \rho }{\alpha \xi^2} \sum_{i=1}^{k} \left(r_i \sqrt{2 s_m} /\sigma \right) \exp\left(- r_i \sqrt{2 s_m} /\sigma\right),
\end{equation}
and conditional on coming from migration the probability the adapted allele 
comes from patch $i$ is 
\begin{align}
  \frac{\left(r_i \sqrt{2 s_m} /\sigma \right) \exp\left(- r_i \sqrt{2 s_m}
    /\sigma\right)} {\sum_{j=1}^{k}  \left(r_i \sqrt{2 s_m} /\sigma \right) \exp\left(- r_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
These specify the transition rates of the process.

@@

If a patch adapts through new mutation or a rare migrant lineage, the
genomic region surrounding the selected site will hitchhike along with
it \citep{maynardsmith1974hitchhiking},
so that at least initially, all adapted individuals within a patch
share a fairly long stretch of haplotype. \gc{Pairs of individuals
  within a patch will share a haplotype of mean genetic length of
  $\approx \approx \log(\rho A s_p)/s_p$ around the selected site,
as long as the patch is reasonably well mixed by dispersal
\citep[otherwise see][]{BartonEtheridge13spatialHH}. }
This association gets slowly whittled down by recombination during interbreeding at the edge of the patch;
but there will always be longer LD nearby to the selected site. 
\plr{Could work out on what time scale it gets whittled down and the shared length at stationarity.}

@@

Above we argued that a good model for this transit is a single Brownian trunk lineage
surrounded by a cloud of close relatives of random size $K$
whose chance of surviving until $t$ generations is $k_e(t) \simeq e^{-\alpha t} / \E[K]$.
Consider a single such family, and let $\tau$ be the (random) time it first hits the new patch,
with $\tau = \infty$ if it never does. The length of the hitchhiking segment depends on the time
spent by this trunk lineage ``in transit'' between the patches, i.e. $\tau$.

By pulling out the constant $1/\E[K]$, we can use standard results on hitting times of $d$-dimensional Brownian motion
that is killed at rate $\alpha$ \citet[][(2.2.0.1 and 4.2.0.1)]{borodin2002handbook} 
In particular, if the patch is circular with radius $w$ and a distance
$R$ from the already adapted patch, then we have the useful result
about the Laplace transform of $\tau$ our transit time
\begin{align}
  \E[e^{-\ell \tau}] &=
    \begin{cases}
      e^{- R \sqrt{2(\alpha+\ell)}/\sigma} \qquad & d=1 \\
      \frac{ K_0( (R+w)\sqrt{2(\alpha+\ell)}/\sigma) }{ K_0( w\sqrt{2(\alpha+\ell)}/\sigma) } \qquad & d=2 
    \end{cases} \label{borodinresult}
    % &= \left( R/w +1 \right)^{-\frac{d-1}{2}} e^{- R \sqrt{2(\alpha+\ell)}/\sigma}  (1+O(1/w)) .
\end{align}
\gc{define $K_0$, bessel? We've used K for family size}
We can find that the probability that a particular
trunk lineageis killed before it reaches the new patch is
$\P\{\tau < \infty\} = \lim_{\ell \to 0} \E \left[\exp(-\ell \tau)
\right]$. We are interested in lineages that
manage to establish on the patch i.e. $\tau< \infty$, and so in one dimension
\begin{equation} \label{eqn:haplen_cdf}
\E[e^{-\ell \tau}|\tau<\infty]  = \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+\alpha)} - \sqrt{2\alpha}\right)}\right\} .
\end{equation}
We can differentiate this to find that our expected transit time is
\begin{equation}
\E[\tau] = (R\sqrt{2\alpha}/\sigma)\times 1/(2\alpha)
\end{equation}
and $\var[\tau] = (R\sqrt{2\alpha}/\sigma) \times 1/(2\alpha)^2$ in
one dimension.

% \simeq (R/w+1)^{-(d-1)/2} \exp(- R \sqrt{2\alpha}/\sigma)$.

% Above, we approximated this by the time $\tau$ it takes the trunk lineage to hit the new patch;
% now that we are studying a successful migrant family means we need to condition on the establishment occurring, i.e.\ on $\tau < \infty$.
% By the above arguments, and equation~\eqref{eqn:estab_prob}, we have that the conditional probability density is
% \begin{align}
%   \P\{\tau\in dt | \tau<\infty \} &= \frac{\P\{\tau \in dt\}}{\P\{\tau<\infty\}} \\
%   &= \frac{R}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{R^2}{2t\sigma^2} -\alpha t + \sqrt{\alpha} \frac{R}{\sigma} \right) .
% \end{align}
% \plr{more dimensions}

Since each generation spent in transit provides an opportunity for recombination,
if recombination is Poisson, the length of the haplotype (in Morgans)
initially shared between populations on each side of the site will be exponentially distributed
with mean $\tau$, so that if $L$ is the length of hitchhiking segment on, say, the right of
the selected locus, then
\begin{equation} \label{eqn:haplen_cdf2}
\P\{L>\ell\} = \E[e^{-\ell \tau}|\tau<\infty] 
%  &= \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+\alpha)} - \sqrt{2\alpha}\right)}\right\} .
\end{equation}
which is our Laplace transform from above that in one dimension
gives us eqn. \eqref{eqn:haplen_cdf}. This implies that in one dimension, if $Y$ is an exponential random variable with rate $R\sqrt{2}/\sigma$,
then $L$ has the same distribution as $(Y + \sqrt{\alpha})^2 - \alpha$,
from which we can obtain that the expected length of shared hitchiking
haplotype is
\begin{equation} \label{eqn:haplotype_length}
\E[L] = \sigma \sqrt{2\alpha}/R + \sigma^2/R^2
\end{equation}
and thar $\var[L] = 2\alpha\sigma^2/R^2 + 4 \sigma^3 \sqrt{2\alpha}/R^3 + 5 \sigma^4 / R^4$.
For two dimensions, asymptotics of Bessel functions show that $L$ has the same tail behavior,
but how other properties might change is unclear. \gc{add some
  sentences of intuition of how parameters play into haplotype length
  here. is it really $\sigma \sqrt{\alpha}$ in \eqref{eqn:haplotype_length} and not $\sqrt{\alpha}/sigma$???}

\gc{Perhaps we should separate the above bit into things that are
  useful about $\tau$ and put those up by  \eqref{borodinresult} and
  things that are useful about $L$ here? We should also turn some of
  this into numbered equations. I tried to do this above, and have the
commented out original below}

% Since each generation spent in transit provides an opportunity for recombination,
% if recombination is Poisson, the length of the hitchhiking segment (in Morgans) on each side of the site will be exponentially distributed
% with mean $\tau$, so that if $L$ is the length of hitchhiking segment on, say, the right of
% the selected locus, then in one dimension
% \begin{align} \label{eqn:haplen_cdf}
% \P\{L>\ell\} &= \E[e^{-\ell \tau}|\tau<\infty] \\
%   &= \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+\alpha)} - \sqrt{2\alpha}\right)}\right\} .
% \end{align}
% found by using eqn. \eqref{borodinresult}. This is also the Laplace transform of $\tau$;
% we can differentiate it to find that $\E[\tau] = (R\sqrt{2\alpha}/\sigma)\times 1/(2\alpha)$
% and $\var[\tau] = (R\sqrt{2\alpha}/\sigma) \times 1/(2\alpha)^2$.
% Furthermore, this implies that if $Y$ is an exponential random variable with rate $R\sqrt{2}/\sigma$,
% then $L$ has the same distribution as $(Y + \sqrt{\alpha})^2 - \alpha$,
% from which we can obtain that
% $\E[L] = \sigma \sqrt{2\alpha}/R + \sigma^2/R^2$ and $\var[L] = 2\alpha\sigma^2/R^2 + 4 \sigma^3 \sqrt{2\alpha}/R^3 + 5 \sigma^4 / R^4$.
% For two dimensions, asymptotics of Bessel functions show that $L$ has the same tail behavior,
% but how other properties might change is unclear.

@@

In the above we already assumed that the inter-patch distance $R$ is large relative to the migration-selection cline width $\sigma/\sqrt{s_m}$;
by taking the large-$R$ limit above we can obtain a more intuitive description of the process
via a central limit for $\tau$.
Define $\eta_\tau$ to be the deviation of $\tau$ from its mean, normalized by its standard deviation,
so that $\tau = R/\sigma(2\alpha)^{-1/2} + \eta_\tau \sqrt{R/\sigma} (2\alpha)^{-3/4}$.
Plugging this into equation~\eqref{eqn:haplen_cdf},
after some algebra (and using that $\sqrt{R+\epsilon} = \sqrt{R} + \frac{\epsilon}{2\sqrt{R}} - \frac{\epsilon^2}{2\sqrt{R}^3} + O(\epsilon^3)$) we obtain that
\begin{align}
  \E[e^{-\ell \eta}] &= \exp\left\{ \frac{\ell^2}{2} + O\left(\frac{1}{R}\right) \right\}.
\end{align}
Since the Laplace transform uniquely determines a distribution \citep{durrett1996probability},
this implies that $\eta_\tau$ converges in distribution to a standard Gaussian random variable as $R \to \infty$.
In other words, we now know that the distribution of $\tau$ about its mean is approximately Gaussian.
Since the standard deviation of $\tau$ is of lower order than its mean,
a similar calculation shows that for large $R$, the haplotype halflength $L$ is approximately exponentially distributed, 
with mean $\sqrt{2\alpha} \sigma/R$.

% \begin{align}
%   \E[e^{-\ell \eta}] &= \E\left[ \exp\left\{-\ell\left(\tau - \frac{R}{\sigma\sqrt{2\alpha}}\right)\frac{(2\alpha)^{3/4}\sqrt{\sigma}}{\sqrt{R}} \right\} \right] \\
%   &= \E\left[ \exp\left\{ - \left(\frac{\ell (2\alpha)^{3/4} \sqrt{\sigma}}{\sqrt{R}}\right) \tau \right\} \right] \exp\left( (2\alpha)^{1/4} \sqrt{\frac{R}{\sigma}} \right) \\
% \end{align}

XXX Check what happens in two dimensions. XXX

@@

>>>>

%%%%%%%%%%%%
\subsection{Multiple patches}
Above we worked with two patches separated by distance $R$;
but these calculations apply as well to multiple patches
in which the $B$ allele can facilitate local adaptation. 
Under our assumptions of mutational exclusion and rapid establishment, 
each patch will adapt through establishment of a single copy of the $B$ allele, 
either by migration or mutation.
If we imagine that the times between successive establishments is the result of many nearly-independent attempts
with small probability of success,
the process of occupation is well approximated by a Markov chain,
whose state records the labeling of patches according to the colonizing allele
(or none, for those that haven't yet adapted).

If not yet adapted, a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$ (equation \eqref{eqn:mutrate}).
Suppose that patches $1, \ldots, k$ are already adapted,
and that these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch.
The total rate of adaptation through migration, from equation \eqref{eqn:migrate}, is
\begin{equation}
  \migrate(r_i) = C \frac{ \rho s_p }{\xi^2 s_m} \sum_{i=1}^{k} \left(r_i \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m} /\sigma\right),
\end{equation}
and if the patch adapts through migration, the probability the adapted allele 
comes from patch $i$ is 
\begin{align}
  \frac{\left(r_i \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m}
  /\sigma\right)} {\sum_{j=1}^{k}  \left(r_i \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
These specify the transition rates of the process.

@@

If a patch adapts through new mutation or a rare migrant lineage, the
genomic region surrounding the selected site will hitchhike along with
it \citep{maynardsmith1974hitchhiking},
so that at least initially, all adapted individuals within a patch
share a fairly long stretch of haplotype. 
Pairs of individuals within a patch will share a haplotype of mean genetic length of
$\approx \log(\rho A s_p)/s_p$ around the selected site, 
as long as the patch is reasonably well mixed by dispersal
\citep[otherwise see][]{barton2013genetic}.
This association gets slowly whittled down by recombination during interbreeding at the edge of the patch;
but there will always be longer LD nearby to the selected site. 
\plr{Could work out on what time scale it gets whittled down and the shared length at stationarity.}

@@

Above we argued that a good model for this transit is a single Brownian trunk lineage
surrounded by a cloud of close relatives of random size $K$
whose chance of surviving until $t$ generations is $k_e(t)$.
Consider a single such family, and let $\tau$ be the (random) time at which it first hits the new patch,
with $\tau = \infty$ if it never does. 
We assume that the length of the hitchhiking segment depends only on 
the time spent by the trunk lineage of the first successful migrant ``in transit'' between the patches,
i.e.\ $\tau$ conditioned on $\tau < \infty$.
In so doing, we neglect 
recombination between B alleles (since they are at low density in transit, we assume this is rare),
and the possibility that more than one migrant family is in transit at once
(so that faster migrants would be more likely to have arrived first);
but these factors should be minor if the rest of our framework fits.

Our first obstacle is that we do not know $k_e(t)$, 
the distribution of the time until a family dies out, exactly.
However, we do know that $k_e(t) \simeq e^{-s_m t} / \E[K]$,
i.e.\ that it has exponential tails.
Since $\E[K] \approx 1/s_m$, it is sensible to make one further approximation,
replacing this random time by an exponential with rate $s_m$.
Once we do this, 
we can use standard results on hitting times of $d$-dimensional Brownian motion
that is killed at rate $s_m$ \citet[][(2.2.0.1 and 4.2.0.1)]{borodin2002handbook} 
In particular, if the patch is circular with radius $w$ and a distance
$R$ from the already adapted patch, then 
we know that the Laplace transform of the transit time $\tau$ is
\begin{align}
  \E[e^{-\ell \tau}] &=
    \begin{cases}
      e^{- R \sqrt{2(s_m+\ell)}/\sigma} \qquad & d=1 \\
      \frac{ K_0( (R+w)\sqrt{2(s_m+\ell)}/\sigma) }{ K_0( w\sqrt{2(s_m+\ell)}/\sigma) } \qquad & d=2  ,
    \end{cases} \label{eqn:borodinresult}
    % &= \left( R/w +1 \right)^{-\frac{d-1}{2}} e^{- R \sqrt{2(s_m+\ell)}/\sigma}  (1+O(1/w)) .
\end{align}
where $K_0$ is a modified Bessel function of the second kind.
We are interested in lineages that manage to reach the patch before being killed,
i.e.\ having $\tau < \infty$,
which occurs with probability
$\P\{\tau < \infty\} = \lim_{\ell \to 0} \E \left[\exp(-\ell \tau) \right]$.

To keep the expressions simple, in the remainder of this section we only compute quantities for one dimension.
By Bayes' rule, the Laplace transform of $\tau$ for a successful lineage is
\begin{equation} \label{eqn:haplen_cdf}
\E[e^{-\ell \tau}|\tau<\infty]  = \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+s_m)} - \sqrt{2s_m}\right)}\right\} .
\end{equation}
This can be differentiated to find that the expected transit time is
\begin{equation} 
  \E[\tau] = (R\sqrt{2s_m}/\sigma)\times 1/(2s_m) \label{eqn:mean_tau}
\end{equation}
and that $\var[\tau] = (R\sqrt{2s_m}/\sigma) \times 1/(2s_m)^2$.

% \simeq (R/w+1)^{-(d-1)/2} \exp(- R \sqrt{2s_m}/\sigma)$.

% Above, we approximated this by the time $\tau$ it takes the trunk lineage to hit the new patch;
% now that we are studying a successful migrant family means we need to condition on the establishment occurring, i.e.\ on $\tau < \infty$.
% By the above arguments, and equation~\eqref{eqn:estab_prob}, we have that the conditional probability density is
% \begin{align}
%   \P\{\tau\in dt | \tau<\infty \} &= \frac{\P\{\tau \in dt\}}{\P\{\tau<\infty\}} \\
%   &= \frac{R}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{R^2}{2t\sigma^2} -s_m t + \sqrt{s_m} \frac{R}{\sigma} \right) .
% \end{align}
% \plr{more dimensions}

Since each generation spent in transit provides an opportunity for recombination,
if recombination is Poisson, the length of the haplotype (in Morgans)
initially shared between populations on each side of the selected locus is exponentially distributed
with mean $\tau$, so that if $L$ is the length of hitchhiking segment on, say, the right of
the selected locus, then
\begin{equation} \label{eqn:haplen_cdf2}
\P\{L>\ell\} = \E[e^{-\ell \tau}|\tau<\infty] ,
%  &= \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+s_m)} - \sqrt{2s_m}\right)}\right\} .
\end{equation}
and so (one minus) equation~\eqref{eqn:haplen_cdf} in fact gives the cumulative distribution function of $L$.
The form of this distribution function implies that if $Y$ is an exponential random variable with rate $R\sqrt{2}/\sigma$,
then $L$ has the same distribution as $(Y + \sqrt{s_m})^2 - s_m$.
Furthermore, the expected length of shared hitchiking haplotype is
\begin{equation} \label{eqn:haplotype_length}
\E[L] = \sigma \sqrt{2s_m}/R + \sigma^2/R^2
\end{equation}
and $\var[L] = 2s_m\sigma^2/R^2 + 4 \sigma^3 \sqrt{2s_m}/R^3 + 5 \sigma^4 / R^4$.
For two dimensions, asymptotics of Bessel functions show that $L$ has the same tail behavior,
but how other properties might change is unclear.

As a general rule of thumb, equation~\eqref{eqn:mean_tau} means that families who successfully establish
move at speed $\sqrt{s_m} \sigma$ towards the new patch 
-- if $s_m$, the strength of the selection against them, is smaller, the need to move quickly is less imperative.
Then, equation~\eqref{eqn:haplotype_length} means that the haplotype length
is roughly the length one would expect given the mean transit time.
Note that we have become used to seeing $\sigma$ divided by $\sqrt{s_m}$, rather than multiplied,
the $\sigma/\sqrt{s_m}$ is a length scale, and to convert it to a time this is divided by $1/s_m$.

\gc{add some
  sentences of intuition of how parameters play into haplotype length
  here. is it really $\sigma \sqrt{s_m}$ in \eqref{eqn:haplotype_length} and not $\sqrt{s_m}/sigma$???}

\gc{Perhaps we should separate the above bit into things that are
  useful about $\tau$ and put those up by  \eqref{eqn:borodinresult} and
  things that are useful about $L$ here? We should also turn some of
  this into numbered equations. I tried to do this above, and have the
commented out original below}

@@

In the above we already assumed that the inter-patch distance $R$ is large relative to the migration-selection cline width $\sigma/\sqrt{s_m}$;
by taking the large-$R$ limit above we can obtain a more intuitive description of the process
via a central limit theorem for $\tau$.
Define $\eta_\tau$ to be the deviation of $\tau$ from its mean, normalized by its standard deviation,
so that $\tau = R/\sigma(2s_m)^{-1/2} + \eta_\tau \sqrt{R/\sigma} (2s_m)^{-3/4}$.
Plugging this into equation~\eqref{eqn:haplen_cdf},
after some algebra (and using that $\sqrt{R+\epsilon} = \sqrt{R} + \frac{\epsilon}{2\sqrt{R}} - \frac{\epsilon^2}{2\sqrt{R}^3} + O(\epsilon^3)$) we obtain that
\begin{align}
  \E[e^{-\ell \eta}] &= \exp\left\{ \frac{\ell^2}{2} + O\left(\frac{1}{R}\right) \right\}.
\end{align}
Since the Laplace transform uniquely determines a distribution \citep{durrett1996probability},
this implies that $\eta_\tau$ converges in distribution to a standard Gaussian random variable as $R \to \infty$.
In other words, we now know that the distribution of $\tau$ about its mean is approximately Gaussian.
Since the standard deviation of $\tau$ is of lower order than its mean,
a similar calculation shows that for large $R$, the haplotype halflength $L$ is approximately exponentially distributed, 
with mean $\sqrt{2s_m} \sigma/R$.

% \begin{align}
%   \E[e^{-\ell \eta}] &= \E\left[ \exp\left\{-\ell\left(\tau - \frac{R}{\sigma\sqrt{2s_m}}\right)\frac{(2s_m)^{3/4}\sqrt{\sigma}}{\sqrt{R}} \right\} \right] \\
%   &= \E\left[ \exp\left\{ - \left(\frac{\ell (2s_m)^{3/4} \sqrt{\sigma}}{\sqrt{R}}\right) \tau \right\} \right] \exp\left( (2s_m)^{1/4} \sqrt{\frac{R}{\sigma}} \right) \\
% \end{align}

\plr{let's leave two dimensions for the revisions. or, as an exercise? =)}

@@

when landscape is not made of discrete (circular) patches
